\paperTitle

### 2014-05-17 doc/icml2014/undirected.tex
Updated figures
Fixed minor typos and stuff in undirected section

% Define model
As before, let $\sG$ be a graphical model with observed variables $X$ and hidden variables $H$
(see \figureref{examples-mrf} for an example).
For each clique $\sC \in \sG$, we have a feature vector $\phi_\sC(\vx_\sC, \vh_\sC)$,
where $\vx_\sC$ and $\vh_\sC$ are the observed and hidden variables in $\sC$, respectively.
Let $\phi(\vx, \vh) = \sum_{\sC \in \sG} \theta^\top \phi_\sC(\vx_\sC,\vh_\sC)$ be the global feature vector.
We consider log-linear models of the following form:
\begin{align}
p_\theta(\vx, \vh) &= \exp\left( \theta^\top \phi(\vx,\vh) - A(\theta) \right),
\end{align}
where $A(\theta) = \log \sum_{\vx, \vh}  \exp( \theta^\top \phi(\vx,\vh) )$ is the log-partition function.

@@

Of course, the log-likelihood is non-concave:
\begin{align}
  L_\text{unsup}(\theta) \eqdef \E_{\vx \sim p^*}[\log \sum_{\vh \in \sH} p_\theta(\vx,\vh)],
\end{align}
where we are assuming for simplicity that we have infinite data drawn from the true distribution $p^*$.
% PL: this doesn't make sense
%While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
%  without loss of generality that $\phi_\sC(x,h)$ are the indicator
%  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
%  statistics for the model, and hence we do not lose any statistical
%  efficiency in using them.}.

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, we would optimize \emph{supervised} likelihood, which is concave:
\begin{align}
\label{eqn:logLinearSupervised}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim p^*}[\log p_\theta(\vx,\vh)] \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\end{align}
While we don't have supervised data,
recall that the method of moments provides the hidden marginals $Z_\sC = \BP(\vh_\sC)$
and conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
Assuming our graphical model is in canonical form (\lemmaref{reduction}),
every observed variable is connected to exactly one hidden variable.
Then $Z_\sC$ and $\mOpp{v}{i}$ directly yield marginals for every clique $\BP(\vx_\sC, \vh_\sC)$.
Given these marginals, we can compute the expected local feature vector:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} \BP(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Note that $\{\mu_\sC\}_{\sC \in \sG}$ are exactly the sufficient statistics of the model (\equationref{logLinearSupervised}),
so we can optimize the supervised likelihood objective without actually having any supervised data!
In the finite data regime, the method of moments yields the estimate
$\hat \mu^\text{mom}_\sC$ which approximates the true $\mu_\sC$.
In supervised learning, we obtain a different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

@@

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{directed}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any bottlenecked clique in the graphical model. 
Consequently, the output of \algorithmref{undirected}
is a consistent estimator.\footnote{Of course, the log-linear model must be identifiable
(it is violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical).
In general, we can identify the parameters up to the kernel of $\nabla^2 A(\theta)$.}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnLogLinear (for undirected models)}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE Conditional moments $\mOpp{v}{i}$ and clique marginals $Z_\sC$.
    \ENSURE Parameters $\theta$
      \STATE Compute expected features $\{\mu_\sC\}_{\sC \in \sG}$ (\equationref{logLinearFeatures}).
      \STATE Optimize \equationref{logLinearSupervised}.
    \end{algorithmic}
\end{algorithm}

@@

Let $\phi_{a,\sN(a)}(a, \sN(a)) = \sum_{\sC \ni a} \phi_\sC(\bx_\sC,\bh_\sC)$
denote the sum over cliques $\sC$ that contain $a$; note that only depends on $a$ and its neighbors $\sN(a)$.
We can write each conditional log-likelihood as:
\begin{align}
p_\theta(a \mid \sN(a)) = \exp(\theta^\top\phi_{a,\sN(a)}(a, \sN(a)) - A_a(\theta; \sN(a))),
\end{align}
where the conditional log-partition function $A_a(\theta; \sN(a)) = \log \sum_{a'} \exp(\theta^\top\phi_{a,\sN(a)}(a', \sN(a)))$
involves summing only over values to a single variable.

If we knew $\mu_{a,\sN(a)} \eqdef \E[\phi_{a,\sN(a)}(a,\sN(a))]$,
then we would be able to optimize the pseudolikelihood objective again without
having access to any labeled data.
%The pseudolikelihood also replaces the log-partition function
%  $A(\theta)$ with a sum over the variables in the neighborhood:
%  $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)}
%  \exp(\theta^\top\phi_{\sN(a)}(a'))$.
Note that $\{a\}\cup\sN(a)$ in general do not form a clique.
Fortunately, \lemmaref{bottleneck-views} holds for all biconnected subgraphs.
We can break $\{a\}\cup\sN(a)$ into a collection of 
  % PL: this is not necessary
  %thus the uniformly bottlenecked assumption alone is not sufficient to
  %guarantee exclusive views for each variable in that subgraph.
%However, when we do have exclusive views for each neighborhood, as in
  \figureref{examples-mrf},
  Therefore, we can adapt our algorithm ($\LearnClique$) to estimate all marginals of the form $Z_{a,\sN(a)} \eqdef \Pr(a,\sN(a))$,
  from which we can compute the features $\mu_{a,\sN(a)}$,
  and optimize \equationref{pseudolikelihood} to obtain a consistent estimate of $\theta$.

>>>>

% Define model
As before, let $\sG$ be a graphical model with observed variables $\bx$ and hidden variables $\bh$
(see \figureref{examples-mrf} for an example).
For each clique $\sC \in \sG$, we have a feature vector $\phi_\sC(\vx_\sC, \vh_\sC)$,
where $\vx_\sC$ and $\vh_\sC$ are the observed and hidden variables in $\sC$, respectively.
Let $\phi(\vx, \vh) = \sum_{\sC \in \sG} \theta^\top \phi_\sC(\vx_\sC,\vh_\sC)$ be the global feature vector.
We consider log-linear models of the following form:
\begin{align*}
p_\theta(\vx, \vh) &= \exp\left( \theta^\top \phi(\vx,\vh) - A(\theta) \right),
\end{align*}
where $A(\theta) = \log \sum_{\vx, \vh}  \exp( \theta^\top \phi(\vx,\vh) )$ is the log-partition function.

@@

Of course, the log-likelihood is non-concave:
\begin{align*}
  L_\text{unsup}(\theta) \eqdef \E_{\vx \sim p^*}[\log \sum_{\vh \in \sH} p_\theta(\vx,\vh)],
\end{align*}
where we are assuming for simplicity that we have infinite data drawn from the true distribution $p^*$.
% PL: this doesn't make sense
%While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
%  without loss of generality that $\phi_\sC(x,h)$ are the indicator
%  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
%  statistics for the model, and hence we do not lose any statistical
%  efficiency in using them.}.

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, we would optimize the \emph{supervised} likelihood, which is concave:
\begin{align}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim p^*}[\log p_\theta(\vx,\vh)] \nonumber \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\label{eqn:logLinearSupervised}
\end{align}
While we don't have supervised data,
recall that the method of moments provides the hidden marginals $Z_\sC = \BP(\vh_\sC)$
and conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
Assuming our graphical model is in canonical form (\lemmaref{reduction}),
every observed variable is connected to exactly one hidden variable.
Then $Z_\sC$ and $\mOpp{v}{i}$ directly yield marginals for every clique $\BP(\vx_\sC, \vh_\sC)$.
Given these marginals, we can compute the expected local feature vector:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} \BP(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Note that $\{\mu_\sC\}_{\sC \in \sG}$ are exactly the sufficient statistics of the model (\equationref{logLinearSupervised}),
so we can optimize the supervised likelihood objective without actually having any supervised data!
In the finite data regime, the method of moments yields the estimate
$\hat \mu^\text{mom}_\sC$ which approximates the true $\mu_\sC$.
In supervised learning, we obtain a different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

@@

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{exclusiveViews}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any bottlenecked clique in the graphical model. 
Consequently, \algorithmref{undirected}
is also a consistent estimator.\footnote{Of course, the log-linear model must be identifiable.
Identifability is violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical.
In general, we can identify the parameters up to the kernel of $\nabla^2 A(\theta)$.}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnParameters (for undirected models)}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE Conditional moments $\mOpp{v}{i}$ and clique marginals $Z_\sC$.
    \ENSURE Parameters $\theta$
      \STATE Compute expected features $\{\mu_\sC\}_{\sC \in \sG}$ (\equationref{logLinearFeatures}).
      \STATE Optimize \equationref{logLinearSupervised}.
    \end{algorithmic}
\end{algorithm}

@@

Let $\phi_{a,\sN(a)}(a, \sN(a)) = \sum_{\sC \ni a} \phi_\sC(\bx_\sC,\bh_\sC)$
denote the sum over cliques $\sC$ that contain $a$; note that $\phi_{a, \sN(a)}$ only depends on $a$ and its neighbors $\sN(a)$.
We can write each conditional log-likelihood as:
\begin{align*}
p_\theta(a \mid \sN(a)) = \exp(\theta^\top\phi_{a,\sN(a)}(a, \sN(a)) - A_a(\theta; \sN(a))),
\end{align*}
where the conditional log-partition function $A_a(\theta; \sN(a)) = \log \sum_{a'} \exp(\theta^\top\phi_{a,\sN(a)}(a', \sN(a)))$
involves marginalizing only over a single variable.

If we knew $\mu_{a,\sN(a)} \eqdef \E[\phi_{a,\sN(a)}(a,\sN(a))]$,
then we would be able to optimize the pseudolikelihood objective again without
having access to any labeled data.
%The pseudolikelihood also replaces the log-partition function
%  $A(\theta)$ with a sum over the variables in the neighborhood:
%  $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)}
%  \exp(\theta^\top\phi_{\sN(a)}(a'))$.
Note that $\{a\}\cup\sN(a)$ in general does not form a clique.
Fortunately, \lemmaref{bottleneck-views} holds for all biconnected subgraphs.
We can break $\{a\}\cup\sN(a)$ into a collection of biconnected subgraphs 
  % PL: this is not necessary
  %thus the uniformly bottlenecked assumption alone is not sufficient to
  %guarantee exclusive views for each variable in that subgraph.
%However, when we do have exclusive views for each neighborhood, as in
 (e.g. \figureref{examples-mrf}).
  Therefore, we can adapt our algorithm ($\LearnMarginals$) to estimate all marginals of the form $Z_{a,\sN(a)} \eqdef \Pr(a,\sN(a))$,
  from which we can compute the features $\mu_{a,\sN(a)}$,
  and optimize \equationref{pseudolikelihood} to obtain a consistent estimate of $\theta$.
